---
title: "Beaks and Flippers"
subtitle: "How Beak Shape and Wing Structure Vary Across Penguin Genera"
author: "Beaky Blinders: Andrew King, Maria Sans-Fuentes, Stephanie Wu"
format: html
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

```{r setup}
# OWNER: Stephanie
library(tidyverse)
library(knitr)

# Color-blind-friendly Okabe-Ito palette (built into R), one color per genus
okabe_ito <- palette.colors(palette = "Okabe-Ito")

genus_cols <- c(
  "Aptenodytes" = unname(okabe_ito["blue"]),
  "Eudyptes"    = unname(okabe_ito["orange"]),
  "Eudyptula"   = unname(okabe_ito["bluishgreen"]),
  "Megadyptes"  = unname(okabe_ito["vermillion"]),
  "Pygoscelis"  = unname(okabe_ito["skyblue"]),
  "Spheniscus"  = unname(okabe_ito["reddishpurple"])
) 

# Define symbols by genus, to support Colour-blind-friendly choices.  
genus_shapes <- c(
  "Aptenodytes" = 16,  # Filled circle
  "Eudyptes" = 17,     # Filled triangle 
  "Eudyptula" = 15,    # Filled square 
  "Megadyptes"  = 18,  # Filled diamond 
  "Pygoscelis" = 3,    # Plus sign (+)
  "Spheniscus" = 8     # Asterisk
)
```

```{r load-data}
# OWNER: Stephanie
penguins <- read_csv("data/many_penguins.csv")

# rename `hand-wing.index` to hwi
# make genus, species and sex factors
```

## Introduction

<!-- OWNER: Stephanie (1-2 paragraphs) -->

Brief introduction to the dataset. You may repeat some of the information about the dataset provided in the introduction to the dataset on the TidyTuesday repository, paraphrasing on your own terms. Imagine that your project is a standalone document and the grader has no prior knowledge of the dataset.  

```{r  data-overview}
# OWNER: Stephanie
# table: number of penguins and species per genus (sample sizes)
# table: number of F, M and U per genus (sex ratio)
# % missing per variable 
```

## Question 1: Do penguin genera differ in beak shape?

### Introduction

<!-- OWNER: Maria (1-2 paragraphs) -->

(1-2 paragraphs): Introduction to the question and what parts of the dataset are necessary to answer the question. Also discuss why you’re interested in this question.

### Approach

<!-- OWNER: Maria (1-2 paragraphs) -->

(1-2 paragraphs): Describe what types of plots you are going to make to address your question. For each plot, provide a clear explanation as to why this plot (e.g. boxplot, barplot, histogram, etc.) is best for providing the information you are asking about. The two plots should be of different types, and at least one of the two plots needs to use either color mapping or
facets. 

- Figure 1: PCA biplot (scatterplot of PC scores, coloured by genus, with loading arrows), faceted into original and size-corrected PCA.
- Figure 2: bar chart of PCA loadings, to show which beak measurement contributes most to each PC.

Also to mention: log-transform and scale; complete cases or imputation; size correction using residuals on tarsus length.

### Analysis   

In this section, provide the code that generates your plots. Use scale functions to provide nice axis labels and guides. You are welcome to use theme functions to customize the appearance of your plot, but you are not required to do so. All plots must be made with {ggplot2}. Do not use base R or lattice plotting functions.

```{r q1-prepare}
# OWNER: Maria
# keep penguins with all four beak measurements (or impute)
# log-transform the four beak measurements
```

```{r q1-pca}
# OWNER: Maria (original PCA)
# prcomp(..., scale. = TRUE) on the log beak measurements

# OWNER: Stephanie (size-corrected PCA)
# residuals of each log beak measurement on log(tarsus_length)
# prcomp(..., scale. = TRUE) on the residuals
```

```{r q1-lda}
# OWNER: Stephanie 
# Linear discriminant analysis: how well can genus be predicted from the beak?

# use the same four log beak measurements (predictors) as the PCA, plus genus 
# (grouping variable to predict)
# MASS package does lda
# table of actual vs predicted genus: table(actual, predicted) - no figure?, 
# we already will have two figures (PCA + loadings)
# % correct overall and per genus
```

```{r fig-q1-biplot, fig.cap="caption for Figure 1"}
# OWNER: Maria
# Figure 1: two biplots side by side (original PCA and size-corrected PCA), PC1 vs PC2
```

```{r fig-q1-loadings,  fig.cap="caption for Figure 2"}
# OWNER: Maria
# Figure 2: bar chart of the PCA loadings
```

### Discussion

<!-- OWNER: Maria and Stephanie (Stephanie: size correction and LDA) -->

(1-3 paragraphs): In the Discussion section, interpret the results of your analysis. Identify any trends revealed (or not revealed) by the plots.Speculate about why the data looks the way it does.

What PC1 and PC2 represent; whether genera group together; which beak measurement contributes most; what changes after size correction; possible reasons (as interpretations); limitations.

## Question 2: Does the hand-wing index still vary among penguin genera?

### Introduction

<!-- OWNER: Andrew (1-2 paragraphs) -->

(1-2 paragraphs): Brief introduction to the dataset. You may repeat some of the information about the dataset provided in the introduction to the dataset on the TidyTuesday repository, paraphrasing on your own terms. Imagine that your project is a standalone document and the grader has no prior knowledge of the dataset.

### Approach

<!-- OWNER: Andrew (1-2 paragraphs) -->

(1-2 paragraphs): Describe what types of plots you are going to make to address your question. For each plot, provide a clear explanation as to why this plot (e.g. boxplot, barplot, histogram, etc.) is best for providing the information you are asking about. The two plots should be of different types, and at least one of the two plots needs to use either color mapping or
facets.   

- Figure 3: boxplot of the hand-wing index by genus, with individual penguins as jittered points coloured by genus and the sample size (n) for each genus.
- Figure 4: mean hand-wing index per genus with 95% confidence intervals (point-and-interval plot), to judge whether genera are similar.

Also mention what test we chose, ANOVA or Kruskal-Wallis, and the effect size as support.

### Analysis

In this section, provide the code that generates your plots. Use scale functions to provide nice axis labels and guides. You are welcome to use theme functions to customize the appearance of your plot, but you are not required to do so. All plots must be made with {ggplot2}. Do not use base R or lattice plotting functions.  


```{r q2-summary}
# OWNER: Andrew
# keep penguins with a hand-wing index value
# summary per genus: n, mean, sd, median, 95% CI of the mean
```

```{r q2-test}
# OWNER: Andrew
# ANOVA or Kruskal-Wallis, and the effect size (eta- or epsilon-squared)
```

```{r fig-q2-boxplot, fig.cap="caption for Figure 3"}
# OWNER: Andrew
# geom_boxplot + geom_jitter, colour and shape = genus, n per genus
```

```{r fig-q2-ci, fig.cap="caption for Figure 4"}
# OWNER: Andrew
# geom_pointrange (mean and 95% CI per genus)
```

```{r q2-explore}
# OWNER: Stephanie
# Exploratory checks for possible reasons why the hand-wing index varies (no figures)
# 1. body size: Spearman correlation between hwi and tarsus_length
# 2. components of the index: mean kipps_distance and wing_length per genus
#    (small table), to see which part drives any differences
# 3. sex: refer to the sex-ratio table from the Introduction; note any genus
#    dominated by one sex
```

### Discussion

<!-- OWNER: Andrew and Stephanie (Stephanie: exploratory checks) -->


(1-3 paragraphs): In the Discussion section, interpret the results of your analysis. Identify any trends revealed (or not revealed) by the plots.Speculate about why the data looks the way it does.   

Do genera differ or are they similar (use the plots, test, effect size and confidence intervals)? Possible reasons, e.g., body size (Stephanie can add a sentence on hand-wing index vs tarsus length) and the adaptation of wings into flippers, as interpretations. Limitations: small samples, penguins of the same species are not independent, no environmental variables.

## References

<!-- OWNER: Stephanie -->

<!-- Everybody review the whole document for consistency -->

## AI Disclosure

<!-- OWNER: All. Each member adds how they used AI (if at all). -->

Indicate whether and how AI tools were used in this project. For each tool, state its name and what it was used for. 

## AI Disclosure

**Maria Sans-Fuentes:** Claude (Anthropic) was used to help choose a colour-blind-friendly palette for our figures. It recommended the Okabe-Ito palette (built into R as `palette.colors(palette = "Okabe-Ito")`), one color per genus. 

**Andrew King:** (state tool and use, or "No AI tools were used.")

**Stephanie Wu:** (state tool and use, or "No AI tools were used.")
